---
title: "Week 5 Lab"
author: "Amy Ly"
date: "4/24/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      error = TRUE, 
                      warning = FALSE)
library(tidyverse)
```

## Notes on Pivoting

# Section 12.3.1

* Real life data is pretty messy.

```{r}
table4a

#pivot to longer table

table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")

```
Remember to use backticks (for variable values/names that exist) when working with tibbles!!

If some name doesn't exist, then use " ". 
```{r}
#does all the renaming 
table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")
```

Combining multiple tibbles
```{r}
tidy4a <- table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")
tidy4b <- table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")
left_join(tidy4a, tidy4b)
```
* Note that pivot_longer() is good for reformating columns names into rows

# Section 12.3.2

* Note that pivot(wider) is good for when you have observations scattered across multiple rows. Then you can group them by column

```{r}
table2

table2 %>%
    pivot_wider(names_from = type, values_from = count)
```
# Section 12.3.3 Exercises

1. Why are Why are pivot_longer() and pivot_wider() not perfectly symmetrical?

pivot_longer() has a names_ptype argument, e.g. names_ptype = list(year = double()). What does it do?

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)

stocks

stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
  pivot_longer(`2015`:`2016`, names_to = "year", values_to = "return")

stocks %>% 
  pivot_wider(names_from = year, values_from = return)


```
Data was lost when we used pivot_wider() and then pivot_longer(). 

Pivot_wider() create column names based on the values in rows, so there would only only 2 distinct values for year and half. 

Then pivot_longer() stacks multiple instances of a value into a column. 

Pivot_long() also changes the data type for the stacked columns. For example, year is a character type now. 

To avoid this issue, one can use the argument names_ptype argument in pivot_longer() to ensure that numeric variables are not changed to character types.

2. Why does this code fail? 

The code fails because there were not backticks used for referencing values that are '1999' and '2000'. As a result, the code tries to select columns 1999 and 2000, which do not exist. 

```{r}
table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")

```

3. What would happen if you widen this table? Why? How could you add a new column to uniquely identify each value?

```{r}
people <- tribble(
  ~name,             ~names,  ~values,
  #-----------------|--------|------
  "Phillip Woods",   "age",       45,
  "Phillip Woods",   "height",   186,
  "Phillip Woods",   "age",       50,
  "Jessica Cordero", "age",       37,
  "Jessica Cordero", "height",   156
)

people

pivot_wider(people, names_from="name", values_from = "values")

pivot_wider(people, names_from="names", values_from = "values")
```

Just using pivot_wider() isn't helpful through name or names isn't helpful. There are very few observations associated with each. 

To make the column wider, I could mutate a dummy column that codes for unique name and names.

```{r}
people2 <- people %>%
  group_by(name, names) %>%
  mutate(obs = row_number())

pivot_wider(people2, names_from="names", values_from = "values")

pivot_wider(people2, names_from="name", values_from = "values")
```
4. Tidy the simple tibble below. Do you need to make it wider or longer? What are the variables?

```{r}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12
)
```

I need to make this table longer. I see that there are multiple responses for 2 columns. 

```{r}
preg_tidy <- preg %>%
  pivot_longer(c(male, female), names_to = "sex", values_to = "count")
preg_tidy

#if I don't want any NA's
preg_tidy <- preg %>%
  pivot_longer(c(male, female), names_to = "sex", values_to = "count", values_drop_na = TRUE)
preg_tidy
```
Now we have 3 different variables: 

* Pregnancy status
* Sex
* Number of people who responded

