---
title: "Chapter 3 Exercises"
author: "Amy Ly"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
#load the tidyverse package
```


## Section 3.2:

Question: Do cars with big engines use more fuel than cars with small engines? 

Exploratory data analysis: 

What is the relationship between engine size and fuel efficiency? 

```{r}
mpg <- get(data(mpg))

head(mpg)
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

```

From the plot, one can see that, on average, that fuel efficiency decreases as engine size increases. 

```{r}
#Alternatively, can also create a ggplot using the following method

ggplot(data = mpg, mapping = (aes(x=displ, y= hwy))) + 
  geom_point() + labs(title = "Fuel Efficiency vs Engine Size", x = "Engine Size", y = "Fuel Efficiency")
```


# 3.2.4 Exercises

$\underline{\text{Problem 1}}$: 

```{r}
ggplot(data=mpg)
```

When I run the above code chunk, I see an empty graph environment. 

$\underline{\text{Problem 2}}$: 

```{r}
dim(mpg)

```
There are 234 rows and 11 columns in the mpg data set.

$\underline{\text{Problem 3}}$: 

Looking at the help page for the mpg dataset, the drv variable describes the type of drive train. 

f = front-wheel drive
r = rear whell drive
4 = 4wd

$\underline{\text{Problem 4}}$:

```{r}
ggplot(data = mpg, mapping = (aes(x=cyl, y= hwy))) + 
  geom_point() + labs(title = "Fuel Efficiency vs Number of Cylinders", x = "Number of Cylinders", y = "Fuel Efficiency")
```

$\underline{\text{Problem 5}}$:

```{r}
ggplot(data = mpg, mapping = (aes(x=drv, y= class))) + 
  geom_point() + labs(title = "Type of Car vs Type of Drive Train", x = "Number of Drive Train", y = "Type of Car")
```

This is not a useful plot because both class and drv are both categorical variables and we shouldn't expect any kind of relationship between categorical variables to be visible on a scatterplot. 

## Section 3.3:

The cars that have higher mileage than expected may be newer models that have been made with more efficient engines or are hybrids. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
# color the points based on the variable "class"
```
Many of the points that fall outside of the linear trend are 2seater cars that are sports cars. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```
Note that each size of the point is associated with a class affiliation. 

```{r}

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
# alpha controls transparency of the points

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
# Shape control shapes of the points
```
Note that ggplot only uses 6 shapes at a time so additional groups will go unplotted, by default. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
#changes color of points to blue
```

Note that setting aesthetic manually goes outside of the aes()

# 3.3.1 Exercises

$\underline{\text{Problem 1}}$: 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

The points in this graph is not blue because the color aesthetic was not specified properly.

The appropriate code is: 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

$\underline{\text{Problem 2}}$:

When you run mpg, you can tell which variables are categorical when you see that the data type for a particular columns is a character type, as indicated by <chr>.

Categorical variables are: 

* manufacturer
* model
* trans
* drv
* fl
* class

Continuous variable are: 

* displ

Integer variables are: 

* year
* cyl
* cty
* hwy

$\underline{\text{Problem 3}}$:

```{r, error=TRUE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = displ))

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = displ))

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = displ))
```
We cannot map a continuous variable to shape. This makes sense since shape would rather have categorical values to map to distinct shapes.

When mapping a continuous variable to color or size, however, we get a gradient of color and sizes. Mapping displ to color is more useful compared to size, since the individual point size is hard to distinguish if a point is within a cluster of other points. 


```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = drv))

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = drv))

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = drv))
```
If I were to map a categorical variable to color, size, or shape then the code works. This is because we have discrete values mapped to a particular category.

$\underline{\text{Problem 4}}$:

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = displ, size = displ))
```

When I map the continuous variable displ to color and size, the code works and a plot is generated. It is redundant, however, since having displ mapped to either one tells the same information. 

$\underline{\text{Problem 5}}$:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) + 
  geom_point(stroke = 0.5, shape = 21)

ggplot(data = mpg, aes(x = displ, y = hwy)) + 
  geom_point(stroke = 2, shape = 21)
```
The stroke aesthetic modifies the width of the border of the points. 

$\underline{\text{Problem 6}}$:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) + 
  geom_point(aes(colour = displ <5))

```

If the aesthetic is mapped to a logical condition, then we have binary data. For example, data that matches the logical condition of having a value less than 5 are "TRUE" and will be colored blue. Otherwise, the data points are colored red. 

## Section 3.4

```{r, error = TRUE}

ggplot(data = mpg) 
+ geom_point(mapping = aes(x = displ, y = hwy))

#this is the wrong position for the "+"

```


